The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Renomeando Sexao 2.1 para "Queue-Based External Events and Stack-Based Internal Events".

\CEU is a synchronous reactive language based on Esterel~\cite{esterel.ieee91} 
with support for multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given time are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
stack-based internal events (Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

@@

\subsection{Stack-Based Internal Events}
\label{sec.ceu.ints}

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions, with
intrinsic queue-based handling, and programs cannot emit them.
%
In contrast, internal events follow a stack-based execution policy, similar to 
subroutine calls in typical programming languages~\cite{ceu.sensys13}.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test.

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick (Esterel is also used 
in hardware design).
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
We believe that for software design, this approach simplifies the reasoning 
about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Furthermore, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of subroutines that supports \CEU's 
reactive statements.

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic semantics to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, also 
disallowing the event-based temporal analysis of \CEU.

@@

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation for trail allocation and scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \emph{C} suitable for 
static scheduling.

>>>>

\CEU is a synchronous reactive language based on Esterel~\cite{esterel.ieee91} 
with support for multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given moment are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

@@

\subsection{Queue-Based External Events and Stack-Based Internal Events}
\label{sec.ceu.ints}

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming 
languages~\cite{ceu.sensys13}.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test (line 11).

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
The event-driven approach is popular in many 
communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
synchronous, even though it is not commonly/formally 
noticed/acknowledged/recognized as such.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Finally, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of coroutines that supports \CEU's 
reactive statements.

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation with respect to trail allocation and 
scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \emph{C} suitable for 
static scheduling.
